We Are All Trying to Describe Something And We Don't Have the Words

### 2023-04-19 outputs/paper/paper.qmd
more edits!!!!

---
title: "We Are All Trying to Describe Something And We Don't Have the Words"
subtitle: "The Jack Antonoffification of Pop Music?"
author: 
  - Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/jack."
date: today
abstract: "In this essay, I will be exploring Jack Antonoff's production style using data from the Spotify API. Drawing from an interview with Caleb Gamman, a guy who went viral for guessing which song on Taylor Swift's Midnights album was produced by Antonoff within seconds, I will delve into the specific characteristics that define an Antonoff-y song, and assess whether or not the audio features defined by Spotify can capture this. I plan to lasso Glen McDonald,the Spotify engineer behind these features, into explaining them to me. There will be a Bayesian inference model. It will be a logistic regression. We do not know whether or not it will be good. Finally, I will armchair a little bit about the cultural significance of Antonoff's music and the reasons behind his ubiquity in the music industry."
format: pdf
number-sections: true
bibliography: bibliography.bib
---

@@

To find an answer to this question, I turned to statistics, people who know more than me about describing music, and the overlord of engineered virality itself -- the Spotify API. There will be a Bayesian inference model. It will be a logistic regression. We do not know whether or not it will be good.

@@

    > I'm literally not an expert and of course I don't actually claim to always be able to be able to 100% identify a Jack Antonoff song. It's not like I've been practicing the way that I can tell if he's produced something. It's just basically if it sounds to me like he's produced it. I know that's annoying I will clarify first, I think it's mostly in the vocal treatment, that's the thing that is the most off-putting to me. But just every decision feels a little bit wrong to me. If I wanted to be cringe I could point out the Juno 6 Bass, the perfectly stacked vocals, the Juno 6 pad, the wide and tight plate reverb, the Juno 6 sword, the goo goo gaga, the man hates treble unless it's just random crap swishing around, everything gets low passed with thin 707 drum samples, thin 808 drum samples, then 909 drum samples, thin Lin drum samples, thin thin drum samples, all the resonances, the harmonics get removed, everything is sort of squished down to the fundamental reduction in timbre, everything exists in a small frequency range, everything becomes a tone, it's an extremely consonant sound with no interplay between the different elements. Now that is a skill to do, and it produces a cleaner sound, and the only problem is I viscerally hate that sound.

@@

    > hing I got into drumming at the age oflike six so maybe as a drummer the drumsamples are of a fence I wasanachronistically raised on cassettes soyou know maybe I like the sound ofaudio and maybe that explains I tend tolike dissonance and in harmonic soundinterplay of elements Distortion noisethe Clashing of elements I imagine I'velistened to more SoundCloud in my lifethan radio so uh maybe I only want tohear amateur internet loser I'veliked over 10 000 songs on my Spotifyaccount uh and I can identify lot ofthem off the bat but I'm not that greatat identifying music generally so it'sjust uh compartmentalized mediaconsumption habit I first did musicproduction in primary school I mademashups and dubstep in it as a lateteenager I extensively used Juno 6emulators to make embarrassing synthwaveadmittedly that's probably not thesubconscious Association that he's goingfor lyrics aren't of much interest to meit's just a sound in the mix and somaybe I don't like the focus on vocalI've previously described it as a fightor flight sort of thing so maybe itliterally is tapping into some memory oftrauma I have an unusually small earcanal so maybe I'm just hearing the highend more acutely I was hired as ateenager to play drums on uh a lot ofJack Antonoff songs for a universityGlee Club performer so maybe I don'tlike it because of the embarrassment ofhaving ever been associated with a Gleecloth I liked the album 1989 but to makethe song out of the woods palette ofwool to me I had to mash it up with M83Midnight City and that got a dmcatakedown so maybe it's a personalVendetta and of course I have known anddo know people who like this music somaybe I secretly hate my friends andfamily but I don't know if any of thattells us anything uh I have a taste inmusic w

@@

    > "i did use the phrase visceral hatred. but i did so only to describe my visceral hatred. i cant stand the sound of it. i hate it. and the way in which i hate it is viscerally. but i didnt go beyond that because it doesnt go beyond that. nothing against the guy, i like taylor swift's stuff. it's just subjective taste...

4.  His tongue in cheek response to some of Twitter's negative response to his video

    > .... \[it's just subjective taste\]. there's just nothing to engage with, so ive had to fabricate a reason for controversy. and what ive come up with is this. \*looks at phone\* the real reason iive been saying for years that i have a visceral reaction to jack antonoffs production style, the REAL reason is this \-- cause during the rollout cycle of the album midnights, i wanted to ascribe taylor swifts authorship to a man and diminish women's voices. so actually, i'm a sexist. so now thats out there, publishhed as facts, in the news. with all the other facts..."

@@

    > "what makes you so special and unique that you can tell the production of a song?" "nothing"

@@

# Numbers?

@@

I pulled the data using the Spotify API, which I accessed using the `spotifyr` package in R [@citeSpotify]. I used the `get_artist_audio_features` and `get_track_audio_analysis functions` function to acquire data on Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, and Bleachers discographies. I threw in HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Von Etten, and Mitski in too, since they've never collaborated with Jack Antonoff but are more or less part of the same pop space.

@@

### Spotify's API

So what do any one these features mean?

The Spotify API returns JSON metadata directly from the Spotify Data Catalogue.

#### Audio Features Dataset

The `get_artist_audio_features` provides a general profile of each song, according to variables developped by Spotify, which makes for simple comparisons across artists. The final audio features in this study data set contains the variables: `artist_name`, `track_name`, `energy`, `danceability`, `key`, `loudness`, `mode`, `speechiness`, `acousticness`, `instrumentalness`, `liveness`, `valence`, `tempo`, `jack`.

\[insert documentation glimpse\]

@@

This is pretty vague. Especially valence -- is the positivity related to lyrics, or something else? I compared the valence scores of classical music tracks to see if this score is based on lyrics, or the audio itself. Jupiter's valence score is 326.78% higher than Mars. This not only indicates the valence is based on audio features other than lyrical content, but that it is in fact capturing a difference between the two songs' profiles.

```{r}

#| echo: false
#| warning: false
library(dplyr)

@@

I looked at the valence measures on Flight of the Bumblebee, a song that's much faster and "upbeat" than Mars and Jupiter. What I found complicated my understanding of valence -- different performances of the same song had different valence scores. Some were arias, played by different instruments. These tend to have higher valence scores. The lowest recorded valence is 0.2410, and the highest is .9230. What gives?

```{r}
sub_bees<- read.csv(here::here("inputs/data/subset_bumblebee.csv"))
sub_bees |>
   select(artist_name, valence, danceability, album_release_year, track_name) |>
   head(17) |>
   knitr::kable(
     col.names = c("Artist Name", 
                   "Valence",
                   "Danceability",
                   "Album Release Year",
                   "Track Name")
  )

@@

Glen told me that `valence` and `danceability` were created by giving thousands of examples to college interns and asking them to tag whether a song was positive or gloomy, or danceable or un-danceable. Variables like energy or instrumentalness, on the other hand, were determined through machine listening techniques. The human subjectivity part only came in when it came to tweaking the training data. My favourite example Glen gave was that the ML kept giving bluegrass songs insanely high `speechiness` ratings. Because there were no banjos in the training data, they were registered as human speech. (This is hilarious.) They had to go back and add more songs with banjos to the training data.

Glen said that the way he's made use of these variables is by combing energy and valence to get a quadrant, which usually turns out "pretty good".

\[discuss the convo we had about how even two humans can have disagreement about a song's positivity, let alone human vs machine. i.e. an elliot smith song where the lyrics are depressing but the melody is happy. or a happy song but you know that the band died in a plane crash yesterday. etc.\]

I thanked Glen for this time, ended the call, and booted up the ❤️statistical computing programming language, R ❤️[@citeR].

```{r}
#| label: fig-quaddanceability2
#| fig-cap: Danceability and Energy Quadrant
#| echo: false
#| warning: false
library(ggplot2)
df |>
  ggplot(aes(x = energy, y = danceability)) + 
    geom_point(aes(color = as.factor(jack)), alpha = 0.3) +
    geom_vline(xintercept = mean(df$energy), color = "gray") +
    geom_hline(yintercept = mean(df$danceability), color = "gray") +
    theme_minimal()

@@

```{r}

#| label: fig-quadvalence2
#| fig-cap: Valence and Energy Quadrant
#| echo: false
#| warning: false

df |>
  ggplot(aes(x = energy, y = valence)) + 
    geom_point(aes(color = as.factor(jack)), alpha = 0.3) +
    geom_vline(xintercept = mean(df$energy), color = "gray") +
    geom_hline(yintercept = mean(df$valence), color = "gray") +
    theme_minimal()

@@

It looks like Jack-produced songs err towards danceability, across the spectrum of energy. It's a bit harder to gauge any trends from the valence quadrant, but it looks like Jack's songs tend to be somewhat on the side of low valence. These quadrants are a start, but they don't suggest any defining Jack-iness characteristic underlying each of those blue points, which are scattered pretty broadly.

Let's throw a logistic regression at the problem.

#library(kableExtra)
#df |>
 # knitr::kable()
```

@@

library(lubridate)
library(tidyverse)
df <- read_csv(here::here("inputs/data/df.csv"))
df$jack <- factor(df$jack)
df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = energy, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Has 'y' in jack?") +
  facet_wrap( ~ artist_name, ncol = 3)

@@

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = danceability, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?") +
  facet_wrap(~ artist_name, ncol = 3)

@@

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = valence, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?") +
  facet_wrap(~ artist_name, ncol = 3)

@@

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = acousticness, x = album_release_date, color = jack)) +
    geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?") +
  facet_wrap(~ artist_name, ncol = 3)

@@

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = instrumentalness, x = album_release_date, color = jack)) +
    geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?") +
  facet_wrap(~ artist_name, ncol = 3)

@@

```{r}

#| label: fig-temp
#| fig-cap: Tempo Variable
#| echo: false
#| warning: false

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = tempo, x = album_release_date, color = jack)) +
    geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?") +
  facet_wrap(~ artist_name, ncol = 3)

@@

I'm gonna fix the axis labels later

```{r}

#| label: fig-quadvalence1
#| fig-cap: Valence and Energy Quadrant 
#| echo: false
#| warning: false

ggplot(data = df, aes(x = energy, y = valence)) + 
  geom_point(aes(color = as.factor(jack)), alpha = 0.3) +
  geom_vline(xintercept = mean(df$energy), color = "gray") +
  geom_hline(yintercept = mean(df$valence), color = "gray") +
  annotate("text", x = mean(df$energy), y = max(df$valence), 
           label = "High Energy, High Valence") +
  annotate("text", x = mean(df$energy), y = min(df$valence), 
           label = "High Energy, Low Valence") +
  annotate("text", x = min(df$energy), y = mean(df$valence), 
           label = "Low Energy, High Valence") +
  annotate("text", x = max(df$energy), y = mean(df$valence), 
           label = "Low Energy, Low Valence")

@@

P(y=1\|x) is the probability of the dependent variable *Y* being 1, given the independent variables *x1*, *x2*, \..., *xn*, and *b0*, *b1*, *b2*, \..., *bn.* The function e\^(-(b0 + b1*x1 + b2*x2 + \... + bn\*xn)), also known as the logit function, is the exponential function of the linear combination of the independent variables and coefficients. βn represents the weights assigned to each independent variable xn. The product of βn and Xn is the contribution of the xn variable to the predicted outcome.

\[Figure X\] gives us an idea of the mean value of each of these features for Jack and non-Jack song. A logistic regression will give us a better idea of how to weigh each of these features in order to predict the outcome, which gives us information about the "research" "question": are these characteristics related to Jack's music? (And then: Which ones? And then, the most elusive question: How?)

@@

```{r}
#|: echo: false
#|: warning: false

data_reduced %>%
  group_by(jack) %>%
  summarise(across(-c(tempo, loudness), mean)) %>%
  pivot_longer(cols = -jack) %>%
  ggplot(aes(value, name, fill = factor(jack))) +
  geom_col(alpha = 0.8, position = "dodge") +
  labs(x = "Mean Value", y = "Feature")

@@

```{r}
#|: echo: false
#|: warning: false

@@

|                            | Actually Positive (1) | Actually Negative (0) |
|----------------------------|-----------------------|-----------------------|
| **Predicted Positive (1)** | 581                   | 25                    |
| **Predicted Negative (0)** | 194                   | 51                    |

@@

```{r}
#jackmatrix
```

```{r}

And here are the different variables' importance on determining whether a song was had any Antonoff involvement.

@@

@@

@@

Table X shows us the 20 songs in Jack Antonoff's discography with the highest danceability scores. Bleachers takes up 35% of this list, and Taylor Swift takes up 40% of this list. St Vincent, The Chicks, Lorde, and Florence and the Machine also haves songs up there.

>>>>

---
title: "We Are All Trying to Describe Something And We Don't Have the Words"
subtitle: "The Jack Antonoffification of Pop Music?"
author: 
  - Michaela Drouillard
thanks: "Code and data are available at: https://github.com/michaeladrouillard/jack."
date: today
abstract: "Using data from the Spotify API, I explored Jack Antonoff's production style and assessed whether Spotify's audio features can capture it. Using a logistic regression model, I found that danceability was the most important feature in predicting Antonoff's involvement. In an interview with Glen McDonald, an engineer at Spotify, I learned that danceability and valence were created by tagging songs as positive or danceable. In the discussion, I do a close-listen to Antonoff's most danceable songs, and argue that what danceability is capturing the visceral quality of Antonoff's music -- which checks out with his critics' visceral reaction to his music."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

To find an answer to this question, I turned to statistics, people who know more than me about describing music, and the overlord of engineered virality itself -- the Spotify API. Using a logistic regression model, I was able to determine that Spotify's `danceability` feature was the most important in predicting whether or not a song had any Antonoff involvement. During interview with Glen McDonald, an engineer at Spotify, I learned that the `valence` and `danceability` variables were created by hiring college interns to tag whether or not songs were positive or danceabile. In the discussion section, I armchair a bit about what I think this `danceability` feature, which has so much uncertainty cooked into it, is really capturing: visceral-ness. Which validates everyone's seemingly visceral reaction to new Antonoff releases, and the popularity of his music.

@@

    > I'm literally not an expert, and of course I don't actually claim to always be able to be able to 100% identify a Jack Antonoff song. It's not like I've been practicing the way that I can tell if he's produced something. It's just basically if it sounds to me like he's produced it. I know that's annoying I will clarify first, I think it's mostly in the vocal treatment, that's the thing that is the most off-putting to me. But just every decision feels a little bit wrong to me. If I wanted to be cringe I could point out the Juno 6 Bass, the perfectly stacked vocals, the Juno 6 pad, the wide and tight plate reverb, the Juno 6 sword, the goo goo gaga, the man hates treble unless it's just random crap swishing around, everything gets low passed with thin 707 drum samples, thin 808 drum samples, then 909 drum samples, thin Lin drum samples, thin thin drum samples, all the resonances, the harmonics get removed, everything is sort of squished down to the fundamental reduction in timbre, everything exists in a small frequency range, everything becomes a tone, it's an extremely consonant sound with no interplay between the different elements. Now that is a skill to do, and it produces a cleaner sound, and the only problem is I viscerally hate that sound.

@@

    > ... I got into drumming at the age of like six, so maybe as a drummer the drum samples are \[offensive?\]. I was anachronistically raised on cassettes so, you know maybe I like the sound of audio and maybe that explains why I tend to like dissonance and inharmonic sound, interplay of elements, distortion noise, the clashing of elements. I imagine I've listened to more SoundCloud in my life than radio so uh, maybe I only want to hear amateur internet losers. I'vel iked over 10 000 songs on my Spotify account and I can identify lot of them off the bat, but I'm not that great at identifying music generally, so it's just a compartmentalized media consumption habit. I first did music production in primary school. I made mashups and dubstep. As a teenager I extensively used Juno 6 emulators to make embarrassing synthwave. \[Admittedly that's probably not the subconscious association that he's going for.\] Lyrics aren't of much interest to me, it's just a sound in the mix, and so maybe I don't like the focus on vocals. I've previously described \[my reaction\] as a fight or flight sort of thing, so maybe it's literally is tapping into some memory of trauma. I have an unusually small ear canal so maybe I'm just hearing the high end more acutely. I was hired as a teenager to play drums on a lot of Jack Antonoff songs for a university glee club so maybe I don't like it because of the embarrassment of having ever been associated with a glee club. I liked the album 1989 but to make the song Out of the Woods palettable to me I had to mash it up with M83Midnight City and that got a DMCA takedown so maybe it's a personal vendetta. And of course I have known andd o know people who like this music so maybe I secretly hate my friends and family but I don't know if any of that tells us anything ...

@@

    > ...I did use the phrase visceral hatred. But I did so only to describe my visceral hatred. I can't stand the sound of it. I hate it. And the way in which I hate it is viscerally. But I didn't go beyond that because it doesn't go beyond that. Nothing against the guy, I like Taylor Swift's stuff. It's just subjective taste...

4.  His tongue in cheek and pretty based response to some of Twitter's negative response to his video

    > .... \[it's just subjective taste\]. There's just nothing to engage with, so I've had to fabricate a reason for controversy. And what I've come up with is this: \**Reads off of phone*\* "The real reason I've been saying for years that I have a visceral reaction to Jack Antonoff's production style, the REAL reason is this: because during the rollout cycle of the album *Midnights*, I wanted to ascribe Taylor Swift's authorship to a man and diminish women's voices." So actually, I'm a sexist. So now that's out there, published as facts, in the news. With all the other facts...

@@

    > .."What makes you so special and unique that you can tell the production of a song?" Nothing. ...

@@

## So

@@

I pulled data on artists who've collaborated with Antonoff using the Spotify API, which I accessed using the `spotifyr` package in R [@citeSpotify]. I used the `get_artist_audio_features` function to acquire data on Lorde, Taylor Swift, St. Vincent, Lana Del Rey, The Chicks, Florence and the Machine, and Bleachers discographies. I threw in HAIM, Marina and the Diamonds, Maggie Rogers, Sharon Von Etten, and Mitski in too, since they've never collaborated with Jack Antonoff but are more or less part of the same pop sphere.

@@

### Audio Features Dataset

The `get_artist_audio_features` provides a profile of each song, according to variables developped by Spotify, which makes for simple comparisons across artists. The final audio features in this study data set contains the variables: `artist_name`, `track_name`, `energy`, `danceability`, `key`, `loudness`, `mode`, `speechiness`, `acousticness`, `instrumentalness`, `liveness`, `valence`, `tempo`, `jack`.

From the `get_artist_audio_features` documentation (@citeSpotifyDocumentation):

> -   **acousticness**: A confidence measure from 0.0 to 1.0 of whether the track is acoustic. 1.0 represents high confidence the track is acoustic.
>
> -   **danceability**: Danceability describes how suitable a track is for dancing based on a combination of musical elements including tempo, rhythm stability, beat strength, and overall regularity. A value of 0.0 is least danceable and 1.0 is most danceable.
>
> -   **energy**: Energy is a measure from 0.0 to 1.0 and represents a perceptual measure of intensity and activity. Typically, energetic tracks feel fast, loud, and noisy. For example, death metal has high energy, while a Bach prelude scores low on the scale. Perceptual features contributing to this attribute include dynamic range, perceived loudness, timbre, onset rate, and general entropy.
>
> -   **instrumentalness**: Predicts whether a track contains no vocals. "Ooh" and "aah" sounds are treated as instrumental in this context. Rap or spoken word tracks are clearly "vocal". The closer the instrumentalness value is to 1.0, the greater likelihood the track contains no vocal content. Values above 0.5 are intended to represent instrumental tracks, but confidence is higher as the value approaches 1.0.
>
> -   **key**: The key the track is in. Integers map to pitches using standard [Pitch Class notation](https://en.wikipedia.org/wiki/Pitch_class). E.g. 0 = C, 1 = C♯/D♭, 2 = D, and so on. If no key was detected, the value is -1.
>
> -   **liveness**: Detects the presence of an audience in the recording. Higher liveness values represent an increased probability that the track was performed live. A value above 0.8 provides strong likelihood that the track is live.
>
> -   **loudness:** The overall loudness of a track in decibels (dB). Loudness values are averaged across the entire track and are useful for comparing relative loudness of tracks. Loudness is the quality of a sound that is the primary psychological correlate of physical strength (amplitude). Values typically range between -60 and 0 db.
>
> -   **mode:** Mode indicates the modality (major or minor) of a track, the type of scale from which its melodic content is derived. Major is represented by 1 and minor is 0.
>
> -   **tempo:** The overall estimated tempo of a track in beats per minute (BPM). In musical terminology, tempo is the speed or pace of a given piece and derives directly from the average beat duration.
>
> -   **valence:** A measure from 0.0 to 1.0 describing the musical positiveness conveyed by a track. Tracks with high valence sound more positive (e.g. happy, cheerful, euphoric), while tracks with low valence sound more negative (e.g. sad, depressed, angry).

@@

This is pretty vague. Especially valence -- is the positivity related to lyrics, or something else? I compared the valence scores of classical music tracks to see if this score is based on lyrics, or the audio itself. In @tbl-Holst, Gustav Holst's Jupiter has a valence score 326.78% higher than Mars. This tracks. It suggests the valence is based on audio features other than lyrical content. It also suggests that it is, in fact, capturing a difference between the two songs' profiles.

```{r}
#| echo: false
#| warning: false
#| label: tbl-Holst
#| tbl-cap: Valence and Danceability Scores for Holst's Mars and Jupiter

library(dplyr)

@@

I looked at the valence measures on Nikolai Rimsky-Korsakov's Flight of the Bumblebee, a song that's much faster and frantic than Mars and Jupiter. What I found complicated my understanding of valence -- different performances of the same song had different valence scores (@tbl-bees). Some were arias, played by different instruments. Arias tended to have higher valence scores. The lowest recorded valence across all performances is 0.2410, and the highest is .9230. What gives?

```{r}
#| echo: false
#| warning: false
#| label: tbl-bees
#| tbl-cap: Valence and Danceability Scores for Every Performance of Flight of the Bumblee on Spotify

sub_bees<- read.csv(here::here("inputs/data/subset_bumblebee.csv"))
sub_bees |>
   select(artist_name, valence, danceability, album_release_year, track_name) |>
   head(17) |>
   knitr::kable(
     col.names = c("Artist Name", 
                   "Valence",
                   "Danceability",
                   "Album Release Year",
                   "Track Name")
  )

@@

Glen told me that `valence` and `danceability` were created by giving thousands of examples to college interns and asking them to tag whether a song was positive or gloomy, or danceable or un-danceable. Variables like energy or instrumentalness, on the other hand, were determined primarily through machine listening techniques. The human subjectivity part only came in when it came to tweaking the training data[^4]:

[^4]: My favourite example Glen gave was that the ML kept giving bluegrass songs insanely high `speechiness` ratings. Because there were no banjos in the training data, they were registered as human speech. (This is hilarious.) They had to go back and add more songs with banjos to the training data.

"You could imagine writing a formula for energy that combines like loudness and tempo and degree of harmonic variation or something. So it was ML, but that one's more like a human helping a machine figure out a formula. Whereas valence is teaching the machine to try to reproduce a purely human thing. Same with danceability. I mean, danceability is whether a human can dance to it.

So like, the machine's not gonna dance, so the machine can't have any opinion on that. The computer could have an opinion on energy. And the computer can definitely have an opinion on loudness. So there's a spectrum from, you know, loudness as purely analytical, and then energy is a little like loudness, with a little more subjectivity. And then danceability and valence are purely subjective."

This put the big range of valence scores for Flight of The Bumblebee into perspective -- what that value is capturing is abstracted away from the actual notes of the song. The valence feature propagates a vibe, so of course this uncertainty makes room for noise and variation.

Glen confirmed that this process didn't know anything about lyrics, which makes valence a particularly difficult variable to built. It could, in theory, pick up on aspects of vocal performance, but it doesn't know anything about languages or words or the meanings of songs. Take an upbeat, happy-sounding Elliot Smith song with devastating lyrics: the machine might register it as happy, where human listeners understand that it's sad. Add to that that two humans might even disagree on the song's valence.

"Plus, we have the confounding factor of like, a song that seems happy today could be sad tomorrow because the singer was killed in a plane crash. The song didn't change, but our world changed and our reactions changed."

All this said, Glen said that the way he's made use of these variables is by combing energy and valence to get a quadrant, which usually works "fairly well". High energy and high valence could be generally happy, cheerful, upbeat. Low energy, low valence could be sad or downbeat. High energy, low valence is sort of angry. Low energy, high valence is serene or calming. I thanked Glen for this time, ended the call, and booted up the ❤️statistical computing programming language, R ❤️[@citeR].

```{r}
#| label: fig-quadvalence
#| fig-cap: Valence and Energy Quadrant
#| echo: false
#| warning: false
library(ggplot2)
df <- read.csv(here::here("inputs/data/df.csv"))

  ggplot(df, aes(x = energy, y = valence)) + 
    geom_point(aes(color = as.factor(jack)), alpha = 0.3) +
    geom_vline(xintercept = mean(df$energy), color = "gray") +
    geom_hline(yintercept = mean(df$valence), color = "gray") +
    theme_minimal()

@@

In @fig-quadvalence, it looks like songs produced by Jack Antonoff are a little more clustered towards the lower end of valence, and across the spectrum of energy, although the variance is pretty big.

```{r}
#| label: fig-quaddanceability2
#| fig-cap: Danceability and Energy Quadrant
#| echo: false
#| warning: false

  ggplot(df, aes(x = energy, y = danceability)) + 
    geom_point(aes(color = as.factor(jack)), alpha = 0.3) +
    geom_vline(xintercept = mean(df$energy), color = "gray") +
    geom_hline(yintercept = mean(df$danceability), color = "gray") +
    theme_minimal()

@@

In @fig-quaddanceability2, it looks like Jack-produced songs err towards higher danceability, across the spectrum of energy.

These quadrants are a start, but they don't suggest any defining Jack-iness characteristic underlying each of those blue points, which are scattered pretty broadly.

By comparing cumulative distribution functions of each variable based on whether or not a song was produced by Antonoff, we can see that some of these variables do capture some sort of underlying Jack-iness (@fig-cdf). For instance, while liveness, instrumentalness, and speechiness are relatively similar, we can see clear differences in danceability, energy, loudness, acousticness, and valence.

```{r}
#| label: fig-cdf
#| fig-cap: Cumulative Distribution Function of Each Variable, 
#| echo: false
#| warning: false
library(tidyverse)
data_reduced <- read.csv(here::here("inputs/data/data_reduced.csv"))

Another way to gauge the Antonoff-ification of pop music could be to look at how artists' sounds have changed over time, and whether or not there is any correlation with Antonoff. Here is where the necessary "correlation is not causation" caveat comes in: suggesting that Antonoff is the *cause* for any change in artists' sounds would be an overstep. These plots are here to get a glimpse at pop's evolution over time, and where Antonoff fits in that picture. Each point represents a song.

@@

library(lubridate)
library(tidyverse)
df <- read_csv(here::here("inputs/data/df.csv"))
df$jack <- factor(df$jack)
df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = energy, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080", "#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Energy") +
  facet_wrap( ~ artist_name, ncol = 3)

@@

In @fig-energy, it looks like there's a downward trend in energy scores in artists who work with Antonoff, but that these downward trends are also somewhat in line with the artists' energy trajectories anyway. Specifically, Lana Del Rey and Taylor Swift. They have enough albums recorded prior to collaborations with Antonoff to be able to observe this. Sharon Von Etten, Maggie Rogers, and Mitski, who have never worked with Antonoff, all have upward trends.

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = danceability, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Danceability") +
  facet_wrap(~ artist_name, ncol = 3)

@@

There aren't any significant correlations in danceability (@fig-danceability) .

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = valence, x = album_release_date, color = jack)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Valence") +
  facet_wrap(~ artist_name, ncol = 3)

@@

In @fig-valence, it looks like Lorde has the only significant change in valence scores, which tracks with my understanding her albums. Melodrama had Green Lights, and Solar Power had a music video of Lorde frolicking on a beach in a yellow bikini with Phoebe Bridgers and Clairo.

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = acousticness, x = album_release_date, color = jack)) +
    geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Acousticness") +
  facet_wrap(~ artist_name, ncol = 3)

@@

Again in @fig-acousticness, Lorde's acoustic Solar Power album jumps out at us. Taylor Swift, Lorde and Haim also have significant-seeming upward trends in acousticness, while Mitski, Sharon Von Etten and Maggie Rogers see decreasing acousticness.

@@

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = instrumentalness, x = album_release_date, color = jack)) +
    geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Instrumentalness") +
  facet_wrap(~ artist_name, ncol = 3)

@@

In @fig-instrumentalness, St. Vincent is the only plot with any significant trend.

```{r}
#| label: fig-tempo
#| fig-cap: Tempo Variable
#| echo: false
#| warning: false

df %>%
  filter(album_release_date_precision == "day") %>%
  mutate(album_release_date = ymd(album_release_date)) %>%
  ggplot(aes(y = tempo, x = album_release_date, color = jack)) +
    geom_point(alpha = 0.3) +
  geom_smooth(method = lm, se = FALSE) +
  theme_minimal() +
  scale_color_manual(values = c("#808080","#FFB6C1")) +
  labs(color = "Has 'y' in jack?", x = "Album Release Date", y = "Tempo") +
  facet_wrap(~ artist_name, ncol = 3)

@@

Finally, in @fig-tempo, there aren't many significant trends, except for Antonoff's collaborations with Lana Del Rey and Taylor Swift.

Now that we understand that there do seem to be some subtle differences in how variables that are and aren't produced by Jack behave, as seen in our cumulative distribution functions, our quadrants, and some minor trends in artists' sounds over time (specifically, `energy`), we can use a logistic regression to see if a model can a) predict whether or not a song was produced by Antonoff based on these variables, and b) tell us which variables were more significant in determining the output.

@@

where $\text{Pr}(y_i=1)$ is the probability of a song being written by Jack, $\beta_0$ is the intercept, and $\beta_1, \ldots, \beta_{10}$ are the coefficients corresponding to the predictor variables.

@fig-means gives us an idea of the mean value of each of these features for Jack and non-Jack song. A logistic regression will give us a better idea of how to weigh each of these features in order to predict the outcome, which gives us information about the "research" "question": are these characteristics related to Jack's music? (And then: Which ones? And then, the most elusive question: How?)

```{r}
#| label: fig-means
#| fig-cap: Comparing the Mean Values of Each Variable
#| echo: false
#| warning: false

data_reduced %>%
  group_by(jack) %>%
  summarise(across(-c(tempo, loudness), mean)) %>%
  pivot_longer(cols = -jack) %>%
  ggplot(aes(value, name, fill = factor(jack))) +
  geom_col(alpha = 0.8, position = "dodge") +
  theme_minimal() +
  labs(fill = "Jack", x = "Mean Value", y = "Feature")

@@

@@

@@

|                        | Actually Positive (1) | Actually Negative (0) |
|------------------------|-----------------------|-----------------------|
| Predicted Positive (1) | 581                   | 25                    |
| Predicted Negative (0) | 194                   | 51                    |

@@

@tbl-importance shows the different variables' importance on determining whether a song was had any Antonoff involvement.

```{r}
#| label: tbl-importance
#| tbl-cap: Importance Scores of Each Variable in Determining Jack-ness
#| echo: false
#| warning: false

@@

```{r}
#| label: fig-importance
#| fig-cap: Importance Scores of Each Variable in Determining Jack-ness
#| echo: false
#| warning: false

@@

```{r}
#| label: tbl-mostdanceable
#| tbl-cap: Top 20 Most Danceable Antonoff-Produced Songs
#| echo: false
#| warning: false

@@

@tbl-mostdanceable shows us the 20 songs in Jack Antonoff's discography with the highest danceability scores. Bleachers takes up 35% of this list, and Taylor Swift takes up 40% of this list. St Vincent, The Chicks, Lorde, and Florence and the Machine also haves songs up there.
